---
title: 'BIOST 561: R Markdown Intro'
author: "David Whitney"
date: "November 2, 2017"
header-includes:
   - \usepackage{graphicx}
output: 
  beamer_presentation:
    theme: "Frankfurt"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

```

## R Markdown: The Basics

The following information is **readily** available if you use Rstudio (e.g. by creating a blank *.Rmd file):

 - Markdown is a simple formatting syntax for authoring HTML, PDF, and MS Word documents. 
 - For more details on using R Markdown see <http://rmarkdown.rstudio.com>.
 - When you click the **Knit** button a document will be generated that includes both content as well as the output of any embedded R code chunks within the document.

## Supported Formats
R Markdown can generate output in several standard formats.

 - **HTML** document, with presentations via **ioslides** or **Slidy**
 - **PDF** document, with presentations via **beamer**
 - **MS Word** document
 
**Note**: we will only scratch the surface today, check out other options at http://rmarkdown.rstudio.com/lesson-9.html.

## Minimalist Document

A `R`Markdown file has extension `*.Rmd` and consists of a YAML metadata header used to specify rendering options and a body of formatted text and code chunks. A simple example is:

```
---
output: html_document
---

That's all, folks!
```

## Rendering Markdown
Though the RStudio GUI is well suited for working with Markdown, you may knit with more refined control using the `render` function:

`rmarkdown::render("example.Rmd", output_format = "html_document")`

## Basic Markdown Syntax

Regardless of your chosen output format, some basic syntax will be useful:

 - Section headers
 - Text emphasis
 - Lists
 - `R` code
 
## Section Headers

To set up different sized header text in your document, use `#` for Header 1, `##` for Header 2, and `###` for Header 3.
 
 - In a presentation, this creates a new slide.

## Text emphasis

 - *Italicize* text via `*Italicize*` or `_Italicize_`.
 - **Bold** text via `**Bold**` or `__Bold__`.

## Unordered Lists

This code

```
* Item 1
* Item 2
    + Item 2a
    + Item 2b
```

Renders these bullets (sub-lists need 1 tab or 4 spaces!)

* Item 1
* Item 2
    + Item 2a
    + Item 2b

## Ordered Lists

This code

```
1. Item 1
2. Item 2
    + Item 2a
    + Item 2b
```

Renders this list (be advised - the bullets may not look great in all templates)

1. Item 1
2. Item 2
  + Item 2a
  + Item 2b

## Inline `R` Code
- To use `R` within a line, use the syntax `` `r '\x60r foo\x60'` ``.
- e.g. `` `r '\x60r round(pi, 5)\x60'` `` renders as `r round(pi, 5)`.
- This can be useful to refer to estimates, confidence intervals, p-values, etc. in the body of an article/homework without worrying about copy errors.

## `R` Code Chunks

- `R` code chunks let you run/render code and results similar to `Sweave` or `knitr`.
- To start a code chunk, use the syntax ```{r chunkName, options}.
- To end the chunk, type ```.

## Example Chunk Output 1

With no options specified, a typical code chunk might look like:

````r
`r ''````{r pressure}
summary(pressure)
```
````

```{r}
summary(pressure)
```

## Example Chunk Output 2

- Want to display the output of a code chunk and the underlying `R` code? 
- Specify the `echo = TRUE` option.

```{r, echo = TRUE}
summary(pressure)
```

## Example Chunk Output 3

- `R` code chunks can also be used to render plots.

```{r, echo = TRUE}
plot(pressure)
```

## Example Chunk Output 4

- Want to display `R` code without evaluating it? 
- Specify `eval = FALSE, echo = TRUE` in the chunk options:

```{r, eval = FALSE, echo = TRUE}
summary(pressure)
```

## Example Chunk Output 5

- For chunks which contain intensive/long computations, you may want to cache the results. (Note: this may not be a _great_ of this!)

````r
`r ''````{r pressure, cache = TRUE}
summary(pressure)
```
````

## More on Chunk Options
For additional information, check out https://yihui.name/knitr/options/

- Note that you can use `knitr::opts_chunk$set(echo = TRUE)` to change the default chunk options.

## Mathematical Symbols/Equations in Markdown
- \LaTeX\'s inline (e.g. `$foo$`) and display (e.g. `$$foo$$`) math modes are supported in Markdown for output to HTML, Word or PDF.

## Tables and Figures in Markdown
- The default `R` output suffices for teaching, but requires cleaning up for assignments, theses, or papers.
- This is when knitting to *.pdf becomes an attractive option.

## Knitting
\begin{figure}
\includegraphics{RMarkdownFlow}
\caption{Image courtesy of http://rmarkdown.rstudio.com/lesson-2.html}
\end{figure}

## \LaTeX\ and Markdown

 - Rendering Markdown as a pdf requires a \LaTeX\ installation.
 - You will additionally need to install Pandoc from http://pandoc.org/
 - With \LaTeX\, many customizations are possible.

## \LaTeX\ Customization, 1

- You can include additional \LaTeX\ commands and content.
- Use the `includes` option as follows to add your favorite style files for the preamble, title/abstract, bibliography, etc...

```
---
title: 'A More Organized Person's Document'
output:
  beamer_presentation:
    includes:
      in_header: header.tex
      before_body: doc_prefix.tex
      after_body: doc_suffix.tex
---
```

## \LaTeX\ Customization, 2

 - If you prefer a self-contained document, you may opt for the `header-includes` option over the modular approach:

```
---
title: 'BIOST 561: R Markdown Intro'
author: "David Whitney"
date: "November 2, 2017"
header-includes:
   - \usepackage{graphicx}
output: 
  beamer_presentation:
    theme: "Frankfurt"
---
```

## Note: \LaTeX in Text
- In Markdown, "`\LaTeX rocks`" renders as "\LaTeX rocks" (no space!).
- Use "`\LaTeX\ rocks`" to render "\LaTeX\ rocks", instead.
- This can be especially important when using new commands.

## Tables

Customization for tables can be carried out using functions in the `knitr` or `xtable` packages from `R`.

````r
`r ''````{r table}
knitr::kable(summary(pressure))
```
````

````r
`r ''````{r table, results='asis'}
xtable::xtable(summary(pressure))
```
````

## Example Output: `knitr`

```{r kable}
knitr::kable(summary(pressure))
```

## Example Output: `xtable`

```{r xtable, results='asis', echo=TRUE}
library(xtable)
print(xtable(summary(pressure)), 
      include.rownames = FALSE, comment = FALSE)
```

## Resources for customizing tables

- `xtable`: https://cran.r-project.org/web/packages/xtable/vignettes/xtableGallery.pdf
- `kable`: https://cran.r-project.org/web/packages/kableExtra/vignettes/awesome_table_in_html.html

## Customizing Figures: Captions

The `fig.cap` option allows you to specify the caption for the figure generated by a given chunk:

````r
`r ''````{r caption, fig.cap="I am the caption"}
plot(pressure)
```
````

## Caption Example

```{r caption, fig.cap='I am the walrus'}
plot(pressure)
```

## Customizing Figures: Size

The `fig.height` and `fig.width` options let you specify the dimensions of your plots:

````r
`r ''````{r caption, fig.height = 4, fig.width = 8}
plot(pressure)
```
````

## Figure Size Example

```{r figsize, fig.height = 4, fig.width = 8}
plot(pressure)
```

## Making it on your own

Altering the default Rmarkdown file each time you write a homework, report, or article would be a pain.

- Fortunately, you don't have to!

## Templates

You can create your own templates which set-up packages, fonts, default chunk options, etc.

- http://rmarkdown.rstudio.com/developer_document_templates.html
- Some packages (e.g `rticles`) provide templates that meet journal requirements or provide other.

## Parameters

You may also set parameters in your document's YAML header

```
---
output: html_document
params:
  date: "2017-11-02"
---
```
or pass new values with the `render` function.

- This creates a read-only list `params` containing the values declared.
- e.g. `params$date` returns `2017-11-02`.

## Child code chunks

To manage large documents, it can be useful to write sections/chapters in separate *.Rmd files and include them as children in the main (parent) file:

````r
`r ''````{r ch1, child = chap1.Rmd}
```
````

- Examples can be found at https://yihui.name/knitr/demo/child/

## Congratulations!

You now have all the tools to start knitting your own documents using `R`Markdown.

## Note

We did not talk much about HTML (or Word, I guess). You can create interactive HTML documents from Rmarkdown with packages such as:

- `shiny`
- `flexdashboard`

## Additional Resources

- RStudio: http://rmarkdown.rstudio.com/
- `xtable` and `knitr` documentation
- Pandoc: http://pandoc.org/
- Google 

## Wrapping up

Any Questions?